Estimate RSV onset and peak timing

### 2023-01-12 RSV_timing_explain/RSV_timing_tutorial.Rmd
add second derivative for onset estimates

## Function to calculate the first and second derivatives
```{r}
deriv <- function(x, y) diff(y) / diff(x) # function to calculate derivative
middle_pts <- function(x) x[-1] - diff(x) / 2 # function to assist derivative calculation
```

```{r}
#install.packages("pspline.inference")
library(pspline.inference)
t <- seq(0.5,12,0.01)
dtime <- seq(0.5,12,0.01)

# we use Iris's package to get the uncertainty interval
# of disease trajectory. 
data=pspline.sample.timeseries(irregular_model, data.frame(time=t), pspline.outbreak.cases, samples=1)

deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),time=c(1:length(diff(t)))) # calculate the first derivative
```

>>>>

## Function to calculate the first and second derivatives

```{r}
deriv <- function(x, y) diff(y) / diff(x) 
# function to calculate derivative

middle_pts <- function(x) x[-1] - diff(x) / 2 
# function to assist derivative calculation
```

```{r, echo=T,message=FALSE}
#install.packages("pspline.inference")
# require did not work for this package
library(pspline.inference)
require(dplyr)

t <- seq(0.5,12,0.01)
dtime <- seq(0.5,12,0.01)
```

## Find uncertainty interval of disease trajectory and onset timing

\scriptsize
```{r}
# we use Iris's package to get the uncertainty interval of disease trajectory. 
cases.samples=pspline.sample.timeseries(
  irregular_model, data.frame(time=t), 
  pspline.outbreak.cases, samples=150)

onset.samples = cases.samples %>% 
  group_by(pspline.sample) %>% # for each sample do the following
  do((function(data){
    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
                            time=c(1:length(diff(t))))
    # calculate the first derivative

    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
      time=c(1:(length(diff(t))-1))) # calculate the second derivative

    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
    # only look at second derivatives in the increasing segment (first derivative>0 )
    second_d_test <- second_d[second_d$time%in%indicator$time,]

    onset = dtime[second_d_test$time[second_d_test$second.deriv==
        max(second_d_test$second.deriv)]] 
# find when the second derivative of  the smooth functions reached its maximum 
    data.frame(
      onset = onset,
      cases = data$cases[which(data$time==onset)]) 
    # find the number of RSV 
    # when the second derivative reach its maximum
  })(.))
```

## Plot the range of the onset on top of the epidemic curves

The dots are what we observed and the grey lines are the fitted epidemic curves. The blue area are the 95% confidence interval of epidemic onsets and the vertical dotted line shows the mean of the epidemic onset estimate.
```{r, message=FALSE,echo=FALSE}
require(ggplot2)
# observation data
Epi_data=data.frame(cases=tail(xt1,12),
                    time=1:12)

names(cases.samples)[3]="cases"
# change the the default column name

ggplot() + 
  geom_line(data=cases.samples,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
  geom_point(data=Epi_data,aes(x=time,y=cases)) + # plot observed case number
geom_vline(xintercept=median(onset.samples$onset),
           linetype="dotted",size=0.2)+ # plot median of the epidemic onset estimate
  annotate("rect", xmin = quantile(onset.samples$onset,0.025),
           xmax = quantile(onset.samples$onset,0.975), 
           ymin = 0, ymax = Inf,
        alpha = .2,fill="blue")+ # show the 95% confidence interval of epidemic onsets
  theme_classic()
```

## Plot the onset estimates of state 2.

The estimates show bimodal distribution because we cut through the epidemic season. To correctly analyze the data, we need to limit the time to after the peak of previous outbreak.

```{r, message=FALSE,echo=FALSE}
irregular_model_2 <- gam(cases ~ s(x=time, k=5, bs="ps"),
                       family=poisson, method="REML", data=data.frame(time=seq(1:12), cases=tail(xt2,12)))

cases.samples_2=pspline.sample.timeseries(
  irregular_model_2, data.frame(time=t), 
  pspline.outbreak.cases, samples=150)

onset.samples_2 = cases.samples_2 %>% 
  group_by(pspline.sample) %>% # for each sample do the following
  do((function(data){
    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
                            time=c(1:length(diff(t))))
    # calculate the first derivative

    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
      time=c(1:(length(diff(t))-1))) # calculate the second derivative

    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
    # only look at second derivatives in the increasing segment (first derivative>0 )
    second_d_test <- second_d[second_d$time%in%indicator$time,]

    onset = dtime[second_d_test$time[second_d_test$second.deriv==
        max(second_d_test$second.deriv)]] 
# find when the second derivative of  the smooth functions reached its maximum 
    data.frame(
      onset = onset,
      cases = data$cases[which(data$time==onset)]) 
    # find the number of RSV 
    # when the second derivative reach its maximum
  })(.))

# observation data
Epi_data_2=data.frame(cases=tail(xt2,12),
                    time=1:12)

names(cases.samples_2)[3]="cases"
# change the the default column name

ggplot() + 
  geom_line(data=cases.samples_2,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
  geom_point(data=Epi_data_2,aes(x=time,y=cases)) + # plot observed case number
geom_vline(xintercept=median(onset.samples_2$onset),
           linetype="dotted",size=0.2)+ # plot mean of the epidemic onset estimate
  geom_vline(xintercept=onset.samples_2$onset,
           linetype="solid",size=0.02)+
  annotate("rect", xmin = quantile(onset.samples_2$onset,0.1),
           xmax = quantile(onset.samples_2$onset,0.9), 
           ymin = 0, ymax = Inf,
        alpha = .2,fill="blue")+ # show the 80% confidence interval of epidemic onsets
  theme_classic()
```

## Retry the analysis for the onset estimates of state 2.

This analysis will work better with weekly data and after the epidemic take off as you will have more data points.

```{r, message=FALSE,echo=FALSE}
irregular_model_2 <- gam(cases ~ s(x=time, k=5, bs="ps"),
                       family=poisson, method="REML", data=data.frame(time=seq(1:11), cases=tail(xt2,11)))

t <- seq(0.5,11,0.01)
dtime <- seq(0.5,11,0.01)

cases.samples_2=pspline.sample.timeseries(
  irregular_model_2, data.frame(time=t), 
  pspline.outbreak.cases, samples=150)

onset.samples_2 = cases.samples_2 %>% 
  group_by(pspline.sample) %>% # for each sample do the following
  do((function(data){
    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
                            time=c(1:length(diff(t))))
    # calculate the first derivative

    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
      time=c(1:(length(diff(t))-1))) # calculate the second derivative

    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
    # only look at second derivatives in the increasing segment (first derivative>0 )
    second_d_test <- second_d[second_d$time%in%indicator$time,]

    onset = dtime[second_d_test$time[second_d_test$second.deriv==
        max(second_d_test$second.deriv)]] 
# find when the second derivative of  the smooth functions reached its maximum 
    data.frame(
      onset = onset,
      cases = data$cases[which(data$time==onset)]) 
    # find the number of RSV 
    # when the second derivative reach its maximum
  })(.))

# observation data
Epi_data_2=data.frame(cases=tail(xt2,11),
                    time=1:11)

names(cases.samples_2)[3]="cases"
# change the the default column name

ggplot() + 
  geom_line(data=cases.samples_2,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
  geom_point(data=Epi_data_2,aes(x=time,y=cases)) + # plot observed case number
geom_vline(xintercept=median(onset.samples_2$onset),
           linetype="dotted",size=0.2)+ # plot median of the epidemic onset estimate
  annotate("rect", xmin = quantile(onset.samples_2$onset,0.025),
           xmax = quantile(onset.samples_2$onset,0.975), 
           ymin = 0, ymax = Inf,
        alpha = .2,fill="blue")+ 
  theme_classic()
```
